Phylogenetic study of Asgard Archaea

### 2020-07-28 MSc_Dissertation_B156476.tex
Added cool methodology_overview; some WIP on introduction

\section{Introduction}

Research questions:  
\begin{enumerate}
\item What are the known patterns of gene expression in high-grade serous ovarian cancer?
\item How can blind-source separation techniques such as independent component analysis and non-negative matrix factorization be applied to transcrptomics studies?
\item How can biological databases be used to link genes into oncogenesis pathways?
\end{enumerate}

@@

A good introduction is found in ref \cite{Stein-OBrien2018} matrix factorization (MF) methods are discussed as a means of extraction the low dimensional structure from a high dimensional ``-omics" datasets, with gene expression analysis being the prime example.  
Conventionally, transcriptomics expression arrays are oriented with genes (or transcripts) as rows, and samples (e.g. patients) in columns.   
A typical expression array might have in the order of tens of thousands of rows and hundreds of columns.  
MF methods reduces this large $M \times N$ (genes $\times$ samples) matrix into two two smaller matrices.
In the terminology of Stein-O'Brien \etal\ these are the $M \times K$ \emph{pattern matrix} and the $K \times N$ \emph{amplitude matrix}.  $K$ is typically small of the order 10, and refers to the number of extracted \emph{factors} (or components).

\begin{figure}[ht]
\def\M{2.2cm}
\def\N{1.3cm}
\def\K{0.3cm}
\def\fudge{-0.8mm}
\centering

\subfloat[Optimising independence of Meta\underline{genes}]{
\begin{tikzpicture}[node distance = 0.25cm]
	\node (X) [draw, fill=green, minimum height=\M, minimum width=\N] {\large X};
	\node [right=of X] (eq) {\large $\approx$};
	\node [right=of eq] (S) [draw, fill=yellow, minimum width=\K, minimum height=\M] {\large S};
	\node [right=of S] (times) {\LARGE$\times$};
	\node [right=of times] (A) [draw, fill=cyan, minimum width=\N, minimum height=\K] {\large A};

@@

	\node [below=\fudge of A] {\tiny{$N$ Samples}};
	\node [left=\fudge of A] {\tiny{$K$}};
	\node [below=of A, text width=2cm, align=center] {\small Metasamples};
\end{tikzpicture}}

\subfloat[Optimising independence of Meta\underline{samples}]{
\begin{tikzpicture}[node distance = 0.25cm]
	\node (X) [draw, fill=green, minimum height=\N, minimum width=\M] {\large X};
	\node [right=of X] (eq) {\large $\approx$};
	\node [right=of eq] (S) [draw, fill=cyan, minimum width=\K, minimum height=\N]{\large S};
	\node [right=of S] (times) {\LARGE$\times$};
	\node [right=of times] (A) [draw, fill=yellow, minimum width=\M, minimum height=\K]{\large A};

@@

	\node [below=\fudge of S] {\tiny{$K$ Metasamples}};
	\node [left=\fudge of S] {\rotatebox{90}{\tiny{$N$ Samples}}};
	\node [below=of S, text width=2cm, align=center] {\small Metasamples};

	\node [below=\fudge of A] {\tiny{$M$ Genes}};
	\node [left=\fudge of A] {\tiny{$K$}};
	\node [below=of A, text width=2cm, align=center] {\small Metagenes};
\end{tikzpicture}}
\caption{Two ways of configuring ICA.  The $S$ matrix is optimised for independence of components, thus configuration (a) optimises for metagenes, while (b) optimises for metasamples. The $X \approx S \times A$ notation is conventional for ICA.}
\label{fig:matrix-factorization}
\end{figure}

There are three matrix factorization methods of potential interest to this project:
\begin{enumerate}
\item Independent component analysis (ICA)
\item Non-negative matrix factorization (NMF)
\item Principal component analysis (PCA)
\end{enumerate}

@@

See figure \ref{fig:matrix-factorization}.

@@

\subsection{Research Questions}
\begin{enumerate}
\item
Which are the influential genes in HGSOC according
to expression data? Do these confirm published
results?
\item
Which of PCA, ICA and NMF is best suited to this
analysis?
\item
For identified genes, what is the underlying biology of
their action?
\item
Can we discriminate subtypes of HGSOC from
expression data?
\item
Is there value in incorporating somatic and germline
genomic data?
\end{enumerate}

@@

\subsubsection{Method outline}
An overview of the methodology adopted in this work can be seen in figure \ref{fig-method-overview}.
In outline our method is follows:
\begin{enumerate}
\item
Identification of a consistent gene list between the TCGA and AOCS datasets.
\item
Unsupervised metagene extraction by matrix factorization methods : NMF, ICA and PCA.
\item
Metagene selection for robustness by k-means clustering of bootstrap resampled factorizations evaluated by silhouette score, based on the TCGA dataset.
\item
Investigation of biological significance of determined factors by gene enrichment analysis against the Gene Ontology (GO).
\item
Computation by least squares optimisation of metasamples associated with with each TCGA derived metagene against TCGA and AOCS datasets.
\item
Alignment of available metadata with computed metasamples, for TCGA and AOCS.
\item 
Survival analysis on TCGA and AOCS to investigate relationship between metasamples and patient survival.
\item
Clustered heatmaps to visualise relationships between metasamples, patients and metadata.
\end{enumerate}

@@

%%%%%%%%%%% RESULTS %%%%%%%%%%%%%
\afterpage{\clearpage}

>>>>

\section{Introduction}
\emph{Say something about ovarian cancer!   An image perhaps.}

@@

A good introduction to field is ref \cite{Stein-OBrien2018}, in which matrix factorization (MF) methods are discussed as a means of extraction the low dimensional structure from a high dimensional ``-omics" datasets, with gene expression analysis being the prime example.

There are three matrix factorization methods of interest to this project:
\begin{enumerate}
\item Independent component analysis (ICA)
\item Non-negative matrix factorization (NMF)
\item Principal component analysis (PCA)
\end{enumerate}

Conventionally, transcriptomics expression arrays are oriented with genes (or transcripts) as rows, and samples (e.g. patients) in columns.   
A typical expression array might have in the order of tens of thousands of rows and hundreds of columns.  
MF methods reduces this large $M \times N$ (genes $\times$ samples) matrix into two two smaller matrices.
In the general terminology of Stein-O'Brien \etal\ these are the $M \times K$ \emph{pattern matrix} and the $K \times N$ \emph{amplitude matrix}.  $K$ is typically small of the order 10, and refers to the number of extracted \emph{factors} (or components). 
Since we are focussing on transriptomics, rather than pattern and amplitude we will use the terms \emph{metagenes} and \emph{metasamples} respectively, which are in common useage.  The symbol conventions for these matrices varies depending whether NMF, ICA or PCA is being discussed.   NMF generally uses $W$ and $H$ respectively while ICA generally uses $S$ and $A$.  PCA is tends to be differently formulated, but $W$ and $T$ sometimes used.   The original matrix to be factorized is variously named $X$ or $V$.

In this work we wish to treat and discuss the three factorization methods in a unified way.  Thus, for the remainder of this dissertation we will adopt the following conventions:
\begin{description}
\item[Expression matrix:] $X$ $(M \times N)$
\item[Metagene matrix:] $W$ $(M \times K)$
\item[Metasample matrix:] $H$ $(K \times N)$
\end{description}
where $N$ is the number of patients (or samples), $M$ is the number of genes and $K$ is the factorization rank, i.e. the number of components or factors.  Thus the factorization is written
\begin{equation}
X \approx W H
\end{equation}
Adding subscripts to indicate the row,column orientation of the matrices:
\begin{equation}
X_{M,N} \approx W_{M,K} H_{K,N}
\end{equation}
as illustrated in figure \ref{fig-matrix-factorization}

\begin{figure}[ht]
\def\M{2.2cm}
\def\N{1.3cm}
\def\K{0.3cm}
\def\fudge{-0.8mm}
\centering
\subfloat[Optimising independence of Meta\underline{genes}]{
\begin{tikzpicture}[node distance = 0.25cm]
	\node (X) [draw, fill=green, minimum height=\M, minimum width=\N] {\large X};
	\node [right=of X] (eq) {\large $\approx$};
	\node [right=of eq] (S) [draw, fill=yellow, minimum width=\K, minimum height=\M] {\large W};
	\node [right=of S] (times) {\LARGE$\times$};
	\node [right=of times] (A) [draw, fill=cyan, minimum width=\N, minimum height=\K] {\large H};

@@

	\node [below=\fudge of A] {\tiny{$N$ Samples}};
	\node [left=\fudge of A] {\tiny{$K$}};
	\node [below=of A, text width=2.5cm, align=center] {\small Metasamples};
\end{tikzpicture}}

\subfloat[Optimising independence of Meta\underline{samples}]{
\begin{tikzpicture}[node distance = 0.25cm]
	\node (X) [draw, fill=green, minimum height=\N, minimum width=\M] {\large X};
	\node [right=of X] (eq) {\large $\approx$};
	\node [right=of eq] (S) [draw, fill=cyan, minimum width=\K, minimum height=\N]{\large W};
	\node [right=of S] (times) {\LARGE$\times$};
	\node [right=of times] (A) [draw, fill=yellow, minimum width=\M, minimum height=\K]{\large H};

@@

	\node [below=\fudge of S] {\tiny{$K$ Metasamples}};
	\node [left=\fudge of S] {\rotatebox{90}{\tiny{$N$ Samples}}};
	\node [below=of S, text width=2.5cm, align=center] {\small Metasamples};

	\node [below=\fudge of A] {\tiny{$M$ Genes}};
	\node [left=\fudge of A] {\tiny{$K$}};
	\node [below=of A, text width=2cm, align=center] {\small Metagenes};
\end{tikzpicture}}
\caption{Two ways of configuring ICA.  The $S$ matrix is optimised for independence of components, thus configuration (a) optimises for metagenes, while (b) optimises for metasamples. The $X \approx S \times A$ notation is conventional for ICA.}
\label{fig-matrix-factorization}
\end{figure}

@@

See figure \ref{fig-matrix-factorization}.

@@

\subsection{Research questions}
The following research goals were set out early in project. 
\begin{enumerate}
\item Which are the influential genes in HGSOC according to gene expression data?   Do these confirm published results?
\item Which of PCA, ICA and NMF is best suited to this analysis?
\item For identified genes, what is the underlying biology?
\item Can we discriminate subtypes of HGSOC from expression data?
\item Is there value in incorporating somatic and germline genomic data?
\end{enumerate}

@@

\subsubsection{Method outline}
An overview of the methodology adopted in this work can be seen in figure \ref{fig-method-overview}.

\begin{figure}
\begin{center}
\includegraphics[scale=0.7, draft=false]{methodlology_overview.pdf}
\end{center}
\caption{Overview of methodology as information flow}
\label{fig-method-overview}
\end{figure}
In outline our method is follows:
\begin{enumerate}
\item
Identification of a consistent gene list between the TCGA and AOCS datasets.
\item
Unsupervised metagene extraction by matrix factorization methods : NMF, ICA and PCA.
\item
Metagene selection for robustness by k-means clustering of bootstrap resampled factorizations evaluated by silhouette score, based on the TCGA dataset.
\item
Investigation of biological significance of determined factors by gene enrichment analysis against the Gene Ontology (GO).
\item
Computation by least squares optimisation of metasamples associated with with each TCGA derived metagene against TCGA and AOCS datasets.
\item
Alignment of available metadata with computed metasamples, for TCGA and AOCS.
\item 
Survival analysis on TCGA and AOCS to investigate relationship between metasamples and patient survival.
\item
Clustered heatmaps to visualise relationships between metasamples, patients and metadata.
\end{enumerate}

@@

%%%%%%%%%%% RESULTS %%%%%%%%%%%%%
% \afterpage{\clearpage}
